---
title: "Personal Project"
author: "Clancy Andrews"
date: "12/7/2021"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Clancy Andrews
# Personal Project
# Fall 2021

## Motivation:

I am using the 2021 Caribou Trail High School Cross Country League rankings as the data set for this project. I chose this data set because I have a strong love for running and competitive racing. In high school I competed in cross country not just at the state level, but at the national level (and had the opportunity to compete internationally). During those four years, I had a strong motivation to improve and compete. The data set I am presenting is the same type of set that is updated throughout the nation on the high school level. Every night, when I got home after a race, I would be able to see my results and track my improvements not just against myself, but other people too. Once high school ended, my love of running still goes on and I am still able to see my younger teammates and competitors improve through the same online database.

## Data Process:

I got the data from Athletic.net, a cross country and track & field database ( https://www.athletic.net/CrossCountry/Division/List.aspx?DivID=61478 ). The database I used does not have a proper method of exporting the data, so I had to manually extract the top 50 individuals in the league. The data that I extracted was fairly small and clean as it was, so there was no need for cleaning. Working with the data, I had a couple of integer value data types that helped with visualization and analysis. The hardest part about the data was that to answer a majority of the questions, I had to convert the times, that were a string data type, to a numeric float data type. That process took a while with some trial and error.

```{r Import Data}
runningdata = read.csv("./CaribouTrailLeagueRankings.csv")
```

```{r Time to Seconds}
for (item in runningdata) {
timeSplit = c(strsplit(runningdata$Time, ':')) #Splits up time string
timeList = as.numeric(unlist(timeSplit)) #Creates a list of numeric values
}
timeMin = timeList[seq(1, length(timeList), 2)] #Gets individual values for minutes
timeSec = timeList[seq(2, length(timeList), 2)] #Gets individual values for seconds
for (item in timeMin) { #Converts minutes to seconds
timeMinToSec = c(timeMin * 60)
}
for (item in timeMinToSec) { #Combines converted minutes and additional seconds
times = c(timeMinToSec + timeSec)
}
#Create new dataframe with added times in seconds column
runningdata1 = runningdata
runningdata1$TimeInSeconds = c(times)
```

## Visualization:

```{r}
plot(x = runningdata1$Rank, y = runningdata1$TimeInSeconds, main = "League Rankings vs Time (Seconds)", xlab = "League Rankings", ylab = "Time (Seconds)")
v = c(runningdata1$Grade)
boxplot(runningdata1$TimeInSeconds ~ runningdata1$Grade, main = "Grade vs Time (Seconds)", xlab = "Grade Level", ylab = "Time (Seconds)")
```
The first image is a scatter plot of the league rankings versus the time in seconds of each competitor. There is a gradual, linear pattern in the middle of the plot but at the top and end ranked runners, the curve becomes more exaggerated. This is due to the top and end ranked runners having different levels of endurance and speed, resulting in a larger gap of time between them and the next runner.

The second image is a box plot of the times for the runners versus the grade levels of the runners. I found this plot interesting. By observation, we see that a majority of the runners in grade 10 and 11 are widely distributed in terms of how fast they can run the race. We can see that grade 9 runners are naturally slower overall (due to body maturity and experience). The most interesting observation is that grade 12 runners are more tightly distributed and are slower overall, compared to the other grade levels. 

## Analysis:

```{r}
library(animation)
kmeans.ani(runningdata1[,c(8,2)], 6)
```

I applied k-means clustering to my data because it can effectively display the cluster of runners based on
there time and grade level. From first glance, you can see that the faster the athlete is, less likely they are
to be younger, i.e., grade 9. You can also see that a majority of the faster clustered runners are grade 11 and 12
and a majority of the runners in the league are grade in grades 10 and 11. A major differentiation between great runners and good runners in the high school setting is if the individual can break the 20 minute mark for the 5000 meters (1200 seconds). When observing the clusters, I have noticed that a majority of the runners who are around the 20 minute mark are older. There is also evidence to show which people are more likely to succeed in the future. This can be seen by the cluster second most from the left.
