---
title: "Personal Project"
author: "Clancy Andrews"
date: "12/7/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Motivation:

I am using the 2021 Caribou Trail High School Cross Country League rankings as the data set for this project. I chose this data set because I have a strong love for running and competitive racing. In high school I competed in cross country not just at the state level, but at the national level (and had the opportunity to compete internationally). During those four years, I had a strong motivation to improve and compete. The data set I am presenting is the same type of set that is updated throughout the nation on the high school level. Every night, when I got home after a race, I would be able to see my results and track my improvements not just against myself, but other people too. Once high school ended, my love of running still goes on and I am still able to see my younger teammates and competitors improve through the same online database.

## Data Process:

I got the data from Athletic.net, a cross country and track & field database ( https://www.athletic.net/CrossCountry/Division/List.aspx?DivID=61478 ). The database I used does not have a proper method of exporting the data, so I had to manually extract the top 50 individuals in the league. The data that I extracted was fairly small and clean as it was, so there was no need for cleaning. Working with the data, I had a couple of integer value data types that helped with visualization and analysis. The hardest part about the data was that to answer a majority of the questions, I had to convert the times, that were a string data type, to a numeric float data type. That process took a while with some trial and error.

```{r Import Data}
runningdata = read.csv("./CaribouTrailLeagueRankings.csv")
```

```{r Time to Seconds}
for (item in runningdata) {
timeSplit = c(strsplit(runningdata$Time, ':')) #Splits up time string
timeList = as.numeric(unlist(timeSplit)) #Creates a list of numeric values
}
timeMin = timeList[seq(1, length(timeList), 2)] #Gets individual values for minutes
timeSec = timeList[seq(2, length(timeList), 2)] #Gets individual values for seconds
for (item in timeMin) { #Converts minutes to seconds
timeMinToSec = c(timeMin * 60)
}
for (item in timeMinToSec) { #Combines converted minutes and additional seconds
times = c(timeMinToSec + timeSec)
}
#Create new dataframe with added times in seconds column
runningdata1 = runningdata
runningdata1$TimeInSeconds = c(times)
```

## Visualization:



## Analysis:

